---
title: "**Impact of Severe Weather Events on public health and economy in the United States**"
author: "Mahmood Salah"
date: "Saturday, May 23, 2015"
output: html_document
---

### **Synopsis**
In this report we aim to describe/analyze  the impact of different weather events on public health and economy based on the storm database collected from the U.S. National Oceanic and Atmospheric Administration's (NOAA) from 1950 - 2011. We will use the estimates of fatalities, injuries, property and crop damage to decide which types of event are most harmful to the population health and economy.From these data, we found that **<%=bigimpact%>**, <%= bigimpact5 %> and <%= bigimpact6 %> are most harmful with respect to population health, while **<%= bigimpact2 %>**, <%= bigimpact7 %> and <%= bigimpact8 %> have the greatest economic consequences.

### **Data Processing**
First we download the storm database from (NOAA) in US that track characteristics of major storms and weather events in the United States, including when and where they occur, as well as estimates of any fatalities, injuries, and property damage.
```{r,echo=TRUE}
# Downloading data if it's not already downloaded.
if(!file.exists("repdata-data-StormData.csv.bz2")) {
        download.file("http://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2",
                      destfile = "repdata-data-StormData.csv.bz2")
}
```

Then we read the generated csv file
```{r,echo=TRUE,cache=TRUE}
dta <- read.csv("data/repdata-data-StormData.csv",sep=",")

```

Subset variables that we need in our questions only
```{r,echo=TRUE}
tidydata <- dta[,c('EVTYPE','FATALITIES','INJURIES', 'PROPDMG', 'PROPDMGEXP', 'CROPDMG', 'CROPDMGEXP')]
```
Convert units (H, K, M, B) to calculate property Damage and crop damage in one column for each type 
```{r,echo=TRUE}
# PROPERTYDMG instead of (PROPDMG and PROPDMGEXP)
tidydata$PROPERTYDMG = 0

tidydata[tidydata$PROPDMGEXP == "H", 'PROPERTYDMG'] = tidydata[tidydata$PROPDMGEXP == "H", 'PROPDMG'] * 10^2
tidydata[tidydata$PROPDMGEXP == "K", 'PROPERTYDMG'] = tidydata[tidydata$PROPDMGEXP == "K", 'PROPDMG'] * 10^3
tidydata[tidydata$PROPDMGEXP == "M", 'PROPERTYDMG'] = tidydata[tidydata$PROPDMGEXP == "M", 'PROPDMG'] * 10^6
tidydata[tidydata$PROPDMGEXP == "B", 'PROPERTYDMG'] = tidydata[tidydata$PROPDMGEXP == "B", 'PROPDMG'] * 10^9
```

```{r,echo=TRUE}
# CROPDAMAGE instead of (CROPDMG and CROPDMGEXP)
tidydata$CROPDAMAGE <- 0 

tidydata[tidydata$CROPDMGEXP == "H", 'CROPDAMAGE'] = tidydata[tidydata$CROPDMGEXP == "H", 'CROPDMG'] * 10^2
tidydata[tidydata$CROPDMGEXP == "K", 'CROPDAMAGE'] = tidydata[tidydata$CROPDMGEXP == "K", 'CROPDMG'] * 10^3
tidydata[tidydata$CROPDMGEXP == "M", 'CROPDAMAGE'] = tidydata[tidydata$CROPDMGEXP == "M", 'CROPDMG'] * 10^6
tidydata[tidydata$CROPDMGEXP == "B", 'CROPDAMAGE'] = tidydata[tidydata$CROPDMGEXP == "B", 'CROPDMG'] * 10^9
```

Convert events which written in lower case to upper case to avoid duplication
```{r,echo=TRUE}
tidydata$EVTYPE <- toupper(tidydata$EVTYPE)
```


### **Results**

### Across the United States,event types that are most harmful with respect to population health
Number of injuries for each event type
```{r,echo=TRUE,cache=TRUE}
evinj <- aggregate(INJURIES~EVTYPE,tidydata,sum)
names(evinj) <- c("Event","Total_injuries")
#sort the table of events by the total number of injuries (decreasing order)
sort.evinj <- evinj[order(evinj$Total_injuries,decreasing = TRUE) , ]
f <- sort.evinj[1,1]
s <- sort.evinj[2,1]
```

Number oF FATALITIES for each event type
```{r,echo=TRUE,cache=TRUE}
evfat <- aggregate(FATALITIES~EVTYPE,tidydata,sum)
names(evfat) <- c("Event_Type","Total_FATALITIES")
#sort the table of events by the total number of FATALITIES (decreasing order)
sort.evfat <- evfat[order(evfat$Total_FATALITIES,decreasing = TRUE) , ]
f1 <- sort.evfat[1,1]
s1 <- sort.evfat[2,1]
```
- The biggest event type lead to FATALITIES is **`r f1`** and the second one is **`r s1`**.
- The biggest event type lead to INJURIES is **`r f`** and the second one is **`r s`**.


Calculate the total number of (INJURIES + FATALITIES) for each event type.
```{r,echo=TRUE}
tidydata$FATorINJ <- tidydata$INJURIES + tidydata$FATALITIES
sumall <- aggregate(FATorINJ~EVTYPE,tidydata,sum)
names(sumall) <- c("Event_Type","FATALITIES_and_INJURIES")

```

In order to answer the qestion, we can make table contain top events have big impact on(FATALITIES and INJURIES).
```{r,echo=TRUE}
#sort the table of events by the total number of FATALITIES and INJURIES (decreasing order)
sort.sumall <- sumall[order(sumall$FATALITIES_and_INJURIES,decreasing = TRUE) , ]

#subset top 10 event types have the biggest impact to use them in darwing the histogram
HEALTHt10 <- sort.sumall[1:10,]
bigimpact <- HEALTHt10[1,1]
```
```{r,echo=FALSE}
#hidden just to appear in the Synopsis
bigimpact <- HEALTHt10[1,1]
bigimpact5 <- HEALTHt10[2,1]
bigimpact6 <- HEALTHt10[3,1]

```
Then we draw histogram between event types and their total number of FATALITIES and INJURIES
```{r,echo=TRUE}
# import ggplot2 library to plot the result

library(ggplot2)

#draw Histogram
ggplot(HEALTHt10, aes(x = Event_Type, y = FATALITIES_and_INJURIES))+
        geom_bar(stat = "identity",fill="indianred2",colour="indianred4")+
        theme(axis.text.x = element_text(angle = 90, hjust = 1,size=12))+
        xlab("Event Type") + ylab("Fatalities and Injuries")+
        ggtitle("Number of Fatalities and Injuries by top 10 Weather Events")

```

- Based on the above histograms, we see that **`r bigimpact `** is the most harmful type to population health.

### Across the United States, types of events that have the greatest economic consequences
Top 10 event types that lead to Property damage.
```{r,echo=TRUE,cache=TRUE}
evprop <- aggregate(PROPERTYDMG~EVTYPE,tidydata,sum)
names(evprop) <- c("Event_Type","Prop_DMG_Cost")

#sort the table of events by the total number of Property damage cost (decreasing order)
sort.evprop <- evprop[order(evprop$Prop_DMG_Cost,decreasing = TRUE) , ]

# Remove rownames to appear ranking from 1 to 10
rownames(sort.evprop) <- NULL

sort.evprop[1:10,]
```


Top 10 event types that lead to Crop damage cost.
```{r,echo=TRUE,cache=TRUE}
evcrop <- aggregate(CROPDAMAGE~EVTYPE,tidydata,sum)
names(evcrop) <- c("Event_Type","Crop_DMG_Cost")
#sort the table of events by the total number of FATALITIES (decreasing order)
sort.evcrop <- evcrop[order(evcrop$Crop_DMG_Cost,decreasing = TRUE) , ]

# Remove rownames to appear ranking from 1 to 10
rownames(sort.evcrop) <- NULL

sort.evcrop[1:10,]
```


Calculate the total cost of Property and Corp damage (CROPDAMAGE + PROPERTYDMG) for each event type.
```{r,echo=TRUE}
tidydata$EC.COSTS <- tidydata$PROPERTYDMG + tidydata$CROPDAMAGE
sumall2 <- aggregate(EC.COSTS~EVTYPE,tidydata,sum)
names(sumall2) <- c("Event_Type","Proppert.Corp_damage")

```

In order to answer the question, we can make table contain top event types that have big impact on Property and Corp damage.
```{r,echo=TRUE}
#sort the table of events by the total  of Property and Corp damage (decreasing order)
sort.sumall2 <- sumall2[order(sumall2$Proppert.Corp_damage,decreasing = TRUE) , ]

#subset top 10 event types have the biggest impact to use them in darwing the histogram
ECONOMYt10 <- sort.sumall2[1:10,]
bigimpact2 <- ECONOMYt10[1,1]
```

```{r,echo=FALSE}
#hidden just to appear in the Synopsis
bigimpact2 <- ECONOMYt10[1,1]
bigimpact7 <- ECONOMYt10[2,1]
bigimpact8 <- ECONOMYt10[3,1]

```

Then we draw a histogram between all event types and their impact on Property and Corp damage.
```{r,echo=TRUE}
ggplot(ECONOMYt10, aes(x = Event_Type, y = Proppert.Corp_damage))+
        geom_bar(stat = "identity",fill="hotpink",colour="hotpink4")+
        theme(axis.text.x = element_text(angle = 90, hjust = 1,size=12))+
        xlab("Event Type") + ylab("Damages in $")+
        ggtitle("Total Property Damage and Crop Damage by \n Severe Weather Events in the U.S. from 1995 - 2011")
```

- Based on the above histograms, we see that **`r bigimpact2`** have the greatest economic consequences.



```{r,echo=FALSE,message=FALSE}
require(knitr)
knit_hooks$set(document = function(x){
  x1 = paste(x, collapse = '\n')
  paste(capture.output(brew::brew(text = x1)), collapse = '\n')
})
```
 <br/> 
  <br/> 
   <br/> 
    <br/> 